---
title: "Reproducible Research: Peer Assessment 1"
author: "Mario Maffioli"
date: "Thursday, March 05, 2015"
output: 
  html_document:
    keep_md: true
---
## load required libraries
```{r}
library(dplyr, warn.conflicts = FALSE )
library(xtable)
library(chron)
library(lattice)

```

## Loading and preprocessing the data

Unzip the activity file and load it into memory
```{r}
rm(list=ls())
unzip("Activity.zip")
activity_raw <- read.csv("activity.csv")


```

Preprocess data sets required for analysis
```{r}

# Summarize raw data to determine steps per date
activity_raw_bydate <- 
    activity_raw %>% group_by(date)%>%
    summarise(steps = sum(steps, na.rm=T))

# Summarize raw data to get the mean of steps per interval
activity_raw_byinterval <- 
    activity_raw %>% group_by(interval)%>%
    summarise(steps_mean = mean(steps, na.rm=T))


# replace NA step values with corresponding interval mean

activity_no_na <- activity_raw

# convert interval ids to factors
activity_no_na$interval  <- as.factor(activity_no_na$interval)

# create named vector to use as look-up table
named_means_vector <- 
    activity_raw_byinterval$steps_mean
names(named_means_vector) <- 
    activity_raw_byinterval$interval

# Extract indices of rows with NA values in steps
indx <- which(is.na(activity_raw$steps))

# Replace NA values with corresponding interval means
activity_no_na[indx,]$steps <- 
        named_means_vector[activity_no_na[indx,]$interval ]

# summarize imputed NA file to get steps per day
activity_noNA_bydate <- 
    activity_no_na %>% group_by(date)%>%
    summarise(steps = sum(steps, na.rm=T))

# create augmented data frame with wend column
activity_no_na_day <- 
    activity_no_na %>%
        mutate(day=is.weekend(as.Date(date)))

# convert Wend  to factor
activity_no_na_day$day  <- as.factor(activity_no_na_day$day)

# label factor levels
levels(activity_no_na_day$day) <- 
    c("weekday", "weekend")

# summarize imputed NA file by interval and week day
activity_no_na_day_byinterval <- 
    activity_no_na_day %>% group_by(interval, day)%>%
    summarise(steps_mean = mean(steps))

# Compute means by date for both raw and imputed NA data sets
means <- c(mean(activity_raw_bydate$steps, na.rm=T), mean(activity_noNA_bydate$steps))

# Compute medians for raw and imputed NA data sets
medians <- c(median(activity_raw_bydate$steps, na.rm=T), median(activity_noNA_bydate$steps))

# build data frame for mean and median comparison
compare_mean_median <- data_frame(mean=means, median=medians)
rownames(compare_mean_median) <- c("with NAs","imputed NAs")

```


## What is mean total number of steps taken per day?

1. Total number of steps taken per day

Table with total number of steps taken per day
```{r xtable, results = 'asis'}

tbl <- xtable(activity_raw_bydate,
           caption="Fig. 1.0\n Steps by date")
align(tbl) <- "|r|r|r|"
print (tbl,
      NA.string="NA",
      type = "html"
      )

```


Create and display histogram
```{r}
par(mfrow = c(1,1))
hist1 <- hist(activity_raw_bydate$steps, breaks=seq(0,25000,2500),xlab = "steps per day", main="Total Steps/Day", col="red", ylim=c(0,25))

```


## What is the average daily activity pattern?

Table showing first 20 rows of average number of steps taken per interval
```{r xtable2, results = 'asis'}

tbl2 <- xtable(head(activity_raw_byinterval,20),
           caption="Fig. 2.0\n Steps by interval")
align(tbl2) <- "|r|c|r|"
print (tbl2,
      NA.string="NA",
      type = "html"
      )

```


Select interval with highest average number of steps
```{r}
max_step_mean <- filter(activity_raw_byinterval, 
       steps_mean == max(steps_mean))
max_step_mean

```

Plot time series of average steps per interval across all days
```{r}
x <- activity_raw_byinterval$interval
y <- activity_raw_byinterval$steps_mean
plot(x,y, type="l", xlab=" 5-minute interval ",
     ylab="average number of steps taken",
     main="Avg. Steps Taken per Interval - All Days", col= "blue", lwd=2)
abline(v = 835, h= 206.1698, lty=2, col="red")
grid(lwd = 2)

```

## Imputing missing values

Find total number of NA's in steps column of raw activity file
```{r}
sum(is.na(activity_raw$steps))

```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

*Strategy*

1. Make a copy of the raw activity data frame.
2. Extract from this copy the indices of the rows with NA values in the 'steps' column.
3. Using the 'activity_raw_byinterval' data frame, create a named vector with the interval means as values and the interval-id as the names. This vector will the used as a look-up table in the next step.
4. Assign the corresponding interval mean to the 'steps' NA value, thus imputing  the interval mean as a replacement for the NA value.


Create a new dataset that is equal to the original dataset but with the missing data filled in. THe table below shows the first 20 rows of data with imputed NAs.

```{r xtable3, results = 'asis'}

tbl3 <- xtable(head(activity_no_na,20),
           caption="Fig. 3.0\n Activity without NAs")
align(tbl3) <- "|r|r|c|r|"
print (tbl3,
      type = "html"
      )

```

Make a histogram of the total number of steps taken each day (with the imputed NA values)
```{r}
op <- par(mfrow = c(1, 2), bg = "lightyellow")

plot(hist1,xlab = "steps per day", main="Total Steps per Day", col="red", ylim=c(0,30))

hist2 <- hist(activity_noNA_bydate$steps, breaks=seq(0,25000,2500),xlab = "steps per day", main="Total Steps/Day (imputeds NA's)", col="cornflowerblue", ylim=c(0,30))

par(op)

```


Compare means and medians with/without NAs
```{r xtable4, results = 'asis'}

tbl4 <- xtable(compare_mean_median,
           caption="Fig. 4.0\n Mean and medians with/without NAs")
align(tbl4) <- "|l|c|c|"
print (tbl4,
      type = "html"
      )

```

## Are there differences in activity patterns between weekdays and weekends?

Panel plot of mean steps per interval for both weekends and weekdays

```{r}
activity_no_na_day_byinterval$interval <- as.integer(activity_no_na_day_byinterval$interval)

# With Latice
xyplot(steps_mean ~ interval | day, data=activity_no_na_day_byinterval, type = "l", layout = c(1, 2),ylab="avg. number steps", main="Average Steps by Five Minute Interval")

```


